### 2014-11-25 papers/paper6.tex
changed chapter headers and added content to paper7
Added content

\chapter*{GameFlow: A Model for Evaluating Player Enjoyment in Games}

>>>>

\chapter*{6) GameFlow: A Model for Evaluating Player Enjoyment in Games}

  \section*{Enjoymen in Media and Games}
  Player enjoyment is the single most important goal for computer games. If Players do not enjoy the game, they will not play the game. Research is often focusing on usability (interface, mechanics and gameplay), rather than enjoyment in games. Many different models have been developed to explain and analyze media enjoyment:

    \begin{itemize}
      \item {\bf Attitude} is a psychological tendency expressed by evaluating a particular entity with some degree of favor or difavor.
      \item {\bf Parasocial interaction} takes place when an audience member develops a parasocial relationship with a media character by talking to the character, imagining, or discussing the life of the character. 
      \item {\bf Transportation theory} suggests that the experience of enjoyment is heighted by an immersion in a narrative world, as well as from the consequence of that immersion. Transportation is a melding of attention, imagery, and feelings.
      \item {\bf The theory of disposition} reletes positive or negative attitudes toward media characters to moral evaluations of their actions, as well as empathy for the main character. 
      \item {\bf Cognition} in relation to media enjoyment involves viewers making judgements on, for instance, the characters ethics, interest, and intelligence. 
    \end{itemize}

  Each theories aims to analyse and understand the enjoyment in terms of one specific aspect or concept. Idividually these theories are fairly narrow. 
  The aim of this study is to develop and validate a model of player enjoyment in games that is based on flow.

  \section*{Enjoyment and Flow}
  A researcher did analyse people who spend large amount of time and effort on activities that are difficult, but provide no external rewards (money, status, etc), such as chess players and rock climbers. Later he analysed ordinary people by asking them how it felt when their lifes were at their fullest and when what they did was most enjoyable. He found that optimal experience, or flow, is the same the world over.

  {\it ``Flow is an experience so gratifying that people are willing to do it for its own sake, with little concern for what they will get out of it, even when it is difficult or dangerous''}.

  Flow experience consists of 8 elements: 
  \begin{enumerate}
    \item a task can be completed;
    \item the ability to concentrate on the task;
    \item that concentration i possible because of the task have clear goals;
    \item that concentration i possible because of the task provides feedback;
    \item the ability to exercise a sense of control over actions;
    \item a deep but effortless involement that removes awareness of the frustration of everyday life;
    \item concern of self disappears, but sense of self emerges stronger afterwards;
    \item the sense of the duration of time is altered
  \end{enumerate}

  \section*{Adapting Flow to Games}

    \subsubsection*{Concentration}
    To be enjoyable, a game has to require concentration, and the player must be able to concentrate on the game. The more concentration a task requires in terms of attention and workload, the more absorbing it will be. Games should grab the players attention quickly and maintain it throughout the game. It is important to increase the players workload but maintain an approperiate level for the players perceptual, cognitive, and memory limits. Also, the players shouldnt be burdened with tasks that dont feel important.

    \subsubsection*{Challenge}
    Challenge is consistently identified as the most important aspect of good game design. Games shouldnt be sufficiently challenging, match the players skill level, vary the level of difficulty, and keep approperiate pace (Not discouraginly hard or boringly easy). Satisfaction in games comes from accomplishing difficult tasks, challenging and surpassing opponents, testing skills, mastering skills, reaching a desired goal, and coping with danger.  (the feeling of triumph).

    \subsubsection*{Player Skills}
    For games to be enjoyable, they must support player skill development and mastery. It is necessary that player develops their skills at playing the game. Intersting tutorials. Rewards are an important part of learning to play a game (learning rights and worngs). Players should also have enough information to start playing the game.

    \subsubsection*{Control}
    In order to experience flow, players must be allowed to exercise a sense of control over their actions. Players should be able to translate their intentions into in-game behavior and feel control of the actual movements of their character. 
    The game should be easy to use, allowing players to start the kind of game they want, turn the game on and off, and save the game in different states. 
    Errors in the game can make players feel they have lost control, especially if the errors or consequence are out of the players control. 
    The changes that players make in the world should be persistent, and be noticeable when players backtrack to where they were before to show that they had impact on the world. 
    The should not be any single optimal strategy for winning and gameplays should be balanced with multiple paths throught the game or ways to win.

    \subsubsection*{Clear Goals}
    Games should present the player with a clear overriding goal early in the game. Each level should have multiple goals.

    \subsubsection*{Feedback}
    Players must recieve feeback at approperiate times. Feedback on direction, distance, progress towards goal. In-game interfaces and sound can deliver feedback. Games should also provide immediate feedback for players actions.

    \subsubsection*{Immersion}
    Players should experience deep but effortless involment in a game. Immersion, engagement, and absoption are concepts that are highlighted. Many game-players devoting entire nights or weekends playing games without being concurrently aware of doing so. Players often have a high level of emotional investment in games due to the time, effort, and attention put into playing. People play games to think thoughts and feel emotions that are related to work, to calm down after a hard day or to escape from everyday worries. Games are often seen as a form of escape from the real world or social norms, or as a way to do things that people otherwise lack the skills, resources, or social permission to.

    \subsubsection*{Social Interaction}
    Games should support and create oppertunities for social interaction. Social interaction is not an element of flow. However, it is clearly a strong element of enjoyment in games, as people play games for social interaction, whether or not they like games or the game they are player. To support social interaction, games should create oppertunities for player competition, cooperation, and connection.

  \section*{Validating the Gameflow Critera}

    \subsubsection*{Evaluation: Warcraft 3}

      \begin{itemize}
        \item {\bf Concentration:} multitude of high-quality stimuli, quickly grabbing the players attention and is held throughout the game, there is not a lot of unimportant tasks, the mission and goal feels important, and the workload is high.
        \item {\bf Challenge:} Not met in skirmish mode since the opponent AI is hard, the difficulty is increasing, can switch race with new challenges. 
        \item {\bf Player Skills:} Players are able to start playing Warcraft 3 immediatley, the tutorial is otional is you are an expert, new objects and features are introduced one at a time, Players are rewarded for their skill development with items, experience, and skills for their heroes. 
        \item {\bf Control:} allows players to feel in control of every aspect of the game, there is no problem with errors or recovering from errors since the game is very polished, the players have freedom and control of their actions and strategies that they use, Players can use their own strategies and choose a suited race for their strategy. 
        \item {\bf Clear Goals:} present both overriding and intermediate goals
        \item {\bf Feedback:} Players are immediatley notified when goals or a mission objectives are completed and can check the status of their goals, subgoals, and completed goals. There is a notification when something new is happening. Players recieve their scores, broken down into heroes, units, and resources. 
        \item {\bf Immersion:} It is easy to become engrossed in Warcraft 3, as there is so much to concentrate on, many tasks to perform, things to monitor, as well as graphics, sound and animation. 
        \item {\bf Social Interaction:} supports competition and cooperation between players, as well as social communities inside and outside the game. 
      \end{itemize}

    \subsubsection*{Evaluation: Lords of EverQuest}

       \begin{itemize}
        \item {\bf Concentration:} numerous stimuli which vary in quality. attractive game world, visually appealing characters, only two or three responses that becomes repetitive. No background story, no motivation or variation. The player is required to focus on unimportant tasks, and the workload is too low. 
        \item {\bf Challenge:} The challenge is below average. Challenge is poor for experienced players, Missions are straightforward, Little variation between missions, Units are very unbalanced. 
        \item {\bf Player Skills:} ok tutorials, familiar but poor interface, Icons are hard to understand because of the details, 
        \item {\bf Control:} do not have full control of several aspects of the game, including units, interface, and gameplay. One good attribute is that the interface is customizeable. If a player fails a mission, the player doesnt find out until the end and need to start over and replay the whole mission. The player does not have freedom in the strategies. Consequently, there is no verety in play styles or strategies available to the player. 
        \item {\bf Clear Goals:} do not present any overriding goal beside that there is a war, no goals are ever really presented. Players are told what to do, but not why. 
        \item {\bf Feedback:} provieds sufficient feedback in general. The player is not given feedback on a mission until the end. Statistics when they level up are not visible. There is no difference in the terrain or visualisation of available places to build. 
        \item {\bf Immersion:} the game is too slow for players to become immersed as there is not enough to concentrate on and not enough challenge. Too much time spent waiting. There is not enough background, character development or storyline for the players to get emotionally involved in the game or connected to characters. 
        \item {\bf Social Interaction:} the game supports competition and cooperation between players. 
      \end{itemize}

    \subsubsection*{Comparisation}

  \section*{Discussion and Conclusions}

### 2014-11-25 papers/paper7.tex
Added content

    \subsubsection*{Third-Party Components}
    It is desireable to be able to leverage third-party products for some of the submodules in a game in order to reduce the workload.

    \subsubsection*{Full-Figure Options}

  \section*{Highly Domain-Specific Requirements}

@@

    \subsubsection*{Crosscutting Concerns}

    \subsubsection*{Depth of Simulation}

    \subsubsection*{Profiling}

    \subsubsection*{Risk}

>>>>

    \subsubsection*{Third-Party Components}
    It is desireable to be able to leverage third-party products for some of the submodules in a game in order to {\bf reduce the workload}. Licensable third-party modules exist for some of the modules, only some of the third-party products have been successful at meeting industry needs. Available products covers: audio, rendering, scene management, collison detection and physics, etc. It can be difficult to use these third-party components; often the programmer must have a lot of experience in the problem domain in order to understand how to interface with the product successfully. Integrating is also a problem because of complex products with dependencies. Since games are CPU-intensive, it will often happen that the third-party component presents a significant {\bf performance bottleneck} for some scenarios - and the programmer must fix these situations or work around them. Often the third-party code fails, and a lot of work is requires, they might as well have written the module from scratch, something you dont want to find out after failing with the licensed code. {\bf cost/benefit analysis}.

    \subsubsection*{Full-Figure Options}
    Instead of licensing components, we can license an entire game engine. If a company is making a game that is not doing anything new technologically, such a license can be a safe decision. If there is new technologically aspects, a game engine can be a poor fir as describes with third-party components.

  \section*{Highly Domain-Specific Requirements}
  Currently there are three levels of programming in games: script code, gameplay code, and engine code. Script and gameplay code control the overall content, rules and high-level behaviour of the game. The engine code provides all the basic mechanisms for simulating IO. Engine code is more complex that gameplay code.

@@

    When writing enige code, you need to be good at engineering and have a lot of domain-specific knowledge. This can be broken into two categories: {\bf Mathematical Knowledge} and {\bf algorithmic knowledge}. When implementing 3D there is required a lot of linear algebra and geometry (2D and 3D). A good engine programmer should have working familiarity with many algorithms. Most algorithms will be mainly focused on one task area, like rendering or physics, but these are algorithms are often very deep and take a while to master.

    \subsubsection*{Crosscutting Concerns}
    To successfully build a engine, it is not enough to understand a lot of math and algorithms. It takes a certain experience and wisdom to choose or dicover algorithms that can be combined into a harmonious whole. When a engine fails, it is often because of the harmony. Figure 3 and 4 are complex systems with a lot of crosscutting concerns. Currently we do not have programming paradigms that help us address this fundamental structual problem.

    \subsubsection*{Depth of Simulation}
    Game code is inherently about simulating some kind of world. In early games, the simulations were primitive. For a while we mainly focused on graphics, which is a simulation of how light behaves in the world. But now we are entering a time when the portions of the simulation governing physics and AI ca be more important to the end users quality and experience. Graphics without good AI is bad! When the graphics looks real, the users expects that the world behave as real as well.

    {\bf Stiffness:} the fact that merely by adjusting constraints, you can cause the simulation to become unstable.
    {\bf Tunneling:} which happens when we integrate across a timestep thats too long, causing us to miss a significant world event. The term tunneling comes from collision detection, where we move entities essentially by teleporting them small distances through space; if we move an entity too quickly, it may pass through a solid object like a wall, unless we take extra steps to detect that situation. Complex problem like $n^{2}$ do not want to be solved in real time.

    \subsubsection*{Profiling}
    Were always trying to push the CPU as far as we can, so profiling is very important. Unfortunatley, there are no good profilers for games. Some vendors of graphics hardware makes some graphics-specific profiling tools, as do the makers of some game consoles. Those tools are also helpful but generally insufficient to get a birds eye view of the system.

    \subsubsection*{Risk}
    Computer games have always evolved toward increased technical complexity to give the players things they never have experienced before. Game developers carry a lot of technical risks as well as game design risk. How will the technology interact with the system and how will the feature feel to the end user? Is it worth all this trouble we are taking implement it?

    {\bf GAMES ARE HARD.}

### 2014-11-25 papers/paper8.tex
changed chapter headers and added content to paper7
Added content

\chapter*{From Visual Simulation to Virtual Reality to Games}

>>>>

\chapter*{8) From Visual Simulation to Virtual Reality to Games}

    \section*{Defining Serious Games}
    People respond differently to the emotionally charged term game depending on wheter they played or did not play video games while growing up.

    {\bf Video Game: a mental contest, played with a computer according to certain rules for amuesements, reaction, or winning a stake.}

    Developing a science of games opens a huge potential for the wider application of games in governmental or corporate training, education, health, public policy, and strategic  communication objectives.

    Electronic Arts defines video games as: ``story, art, and software''. 
    Serious games have more than just a story, art and software. They involve pedagogy: activities that educate or instruct. This addition makes games serious. Pedagogy must be subordinate to story - the entertainment component comes first.

    \section*{Creating a Science og Games}
    The development and wide release of the Americas Army game began an revolution in thinking about the potential role of video games for nonentertainment domains. A study of the game was conducted to see if  it could be used for training. 
    Example of use: when new recruits had trouble with the rifle or the obstacles course, his team had those recruits play ``Americas Army'' and required them to complete those levels in the game. When the reqruits did so, they then went back to the range and usually passed the range tests. This made us wonder what other applications would benefit from game-based instruction.
    Mothers complained about their sons playing the game all the time and their kids knew everything about the army. These comments led the researchers to wonder how much of science and math education could be taught via games (learning from other than formal teaching). This is called ``first-person education''.

    \section*{Game Production Challenges}
    First-person shooter games are recently the only games that have reuseable engines. Game engines are costly to develop! The demand for better computer characters and story increases with the complexity of visual displays and more complex games. Game play innovation is becomming a competitive necessity. The big hits have been sports, first-person shooters, adventure, and sims-type games. For the game industry to continue to grow, additional genres must become more sophisticated with better backstories and better game-play technologies.

    \section*{Games Reseach Agenda}
    To influence the future of both serious and entertainment games, developers must create a research agenda that transforms the game production process form a handcrafted, labor-intensive effort to one shorter, more predictable production timelines that still manages to provide innovations and increased complexity. This reaseach agenda have three components: infrastructure, cognitive game design, and immersion.

      \subsubsection*{Infrastructure}
      Underlying software and hardware necessary for developing interactive games include massiveley multiplayer online game architectures, game engines and tools, streaming media, next-greneration consoles, and wireless and mobile devices. Game engines and tools will be vital to researchers bent on attacking the lack-of-reuse problem in gaming. This can also move games to the government and corporate sectors. Developers need an open source game engine that includes a development toolset as wideley available and utilized as Linux. Streaming media will play prominent role in the delivery of dynamic content to PC -based games and mobile devices.

      \subsubsection*{Cognitive game design}
      Taking a cognitive approach to game developemt will give developers the tools to create theories and methods for:

        \begin{itemize}
          \item modeling and simulating computer characters, story, and human emotion
          \item analyzing large-scale games
          \item innovating new game genres and play styles
          \item integrating pedagogy with story  in the interactive game medium 
        \end{itemize}

      The Sims for a serious purpose, such as training aid for nursing. Developers can use this approach to model and simulate hospital operations in game form, providing an immersive experience for the nurse trainee.

      It must be determined what happens during game play and how the experiences affects the players. The game industry has already witnessed tha failure of edutainment, an awkward combination of educational software sprinkled with game-like interfaces and cut dialog. This shows that story must come firstand that research must focus on combining instruction with story creation and the game development process.

      \subsubsection*{Immersion}
      Creating technologies that engage the game players mind via:
        \begin{itemize}
          \item computer graphics, sound, and haptics
          \item affective computing - sensing human state and emotion
          \item advanced user interfaces
        \end{itemize}
      In the next years, low cost sensors will become available that measures the players emotional state and provides this information as input when running the game. Game designers must understand these things if they are to engineer and implement these capabilities so that these features behave predictabily and reabily.
      Advanced user interfaces will become key as computing moves from the standard desktop PC to mobile platforms. Much can be gained by studying how the game industry has developed almost universal interfaces that let gamers transision seemlessly from one game to another. To make significant progress in deploying serious games, developers must understand interfaces from the game perspective.

    \section*{Serious Games}
    Applying games and simulation technology to nonentertainment domains results in serious games.

    \section*{Gamepipe Laboratory}
    The first serious games must be constructed in carefully controlled university environments. The pipeline laboratorys mission is to research and develop interactive game technologies that dramatically shorten the production timeline, provide supporting technologies for increasing the complexity and innovation in produced games, and enable the creation of serious and entertainment games for government and corporate sponsors. The GamePipe Laboratory performs research and development required for the next generation game production. It produces serious games and game prototypes as sponsored projects. It also links with, and develops educational programs in game developement and game-related research and build cross-campus interdisiplinary teams to solve next-generation game technology and development problems.
